# Data Visualization

## `ggplot2 ()` Package - Data Visualization Tool

The package `ggplot2 ()` is a very powerful tool for data visualization. 

Please use the following website to develop this chapter - 
<https://r-graphics.org/index.html>

## Scatter Plot 

In scatter Plot, we use to continuous variables in x axis and y axis. The following code is run to create a scatter plot of two variables called `time_hour` and `total_delay`. See Figure \@ref(fig:scatterplot1). 

```{r}
library(nycflights13)
library(tidyverse)
flights <- flights %>% 
  mutate(total_delay = dep_delay + arr_delay)
```


```{r scatterplot1, fig.cap= "A Scatterplot of Total Delay and Month"}
flights %>% 
  ggplot(mapping = aes( x = time_hour, y = total_delay))+
  geom_point()
```

However, we can also use a thrid varible in scatter plot. See Figure \@ref(fig:scatterplot2). Also see Figure \@ref(fig:scatterplot3). 

```{r scatterplot2, fig.cap= "A Scatterplot of Total Delay and Month in Departing Airports in New York"}
flights %>% 
  ggplot(mapping = aes( x = time_hour, y = total_delay))+
  geom_point()+
  facet_wrap(~origin)

```


```{r scatterplot3, fig.cap= "A Scatterplot of Total Delay and Month Using Origin as Third Variable"}
flights %>% 
  ggplot(mapping = aes( time_hour, total_delay, color = origin))+
  geom_point()
```

## Line Plot  

In line plot, we draw line for the points of two continuous variables. See Figure \@ref(fig:lineplot1).

```{r lineplot1, fig.cap= "A lineplot of Average Delay and Month"}
flights %>% 
  group_by(month) %>% 
  summarize(avg_delay = mean (total_delay, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = month, y = avg_delay)) + 
  geom_smooth()
```

Like scatter plot, a third variable can also be added to the line plot. See Figure \@ref(fig:lineplot2). Also see Figure \@ref(fig:lineplot3). 

```{r lineplot2, fig.cap= "A lineplot of Average Delay and Month Using Origin as Third Variable" }
flights %>% 
  ggplot(mapping = aes(x = month, y = total_delay, color = origin)) + 
  geom_smooth()
```


```{r lineplot3, fig.cap= "A lineplot of Total Delay and Day of the Week Using Origin as Third Variable"}
flights %>% 
  ggplot(mapping = aes(x = day, y = total_delay, color = origin)) + 
  geom_smooth()
```

## Bar Plot  

We can also create bar diagram. See Figure \@ref(fig:barplot1).

```{r barplot1, fig.cap= "A barplot of Number of Flights from Departing Airports in New York"}
ggplot(flights, aes(origin))+
  geom_bar()
```

We can also include a second variable in bar diagram. Please see Figure \@ref(fig:barplot2). Also see Figure \@ref(fig:barplot3).

```{r barplot2, fig.cap= "A barplot of Number of Flights from Departing Airports in New York in Different Months"}
ggplot(flights, aes(origin))+
  geom_bar(aes(fill = as.character(month)))
```

```{r barplot3, fig.cap= "A barplot of Number of Flights from Departing Airports in New York by Different Carriers"}
ggplot(flights, aes(origin))+
  geom_bar(aes(fill = carrier))
```

From the graph, we can figure out which carriers use which airport most. 

## Box Plot 

We can also create boxplot using `ggplot2`. Please see Figure \@ref(fig:boxplot1).

```{r boxplot1, fig.cap= "A boxplot of Distance from Departing Airports in New York"}
ggplot(flights, aes ( origin,distance))+
  geom_boxplot()
```

It is evident that long distance flights use JFK because it has the highest distance. A thrid variable also can be included in box plot. Please see Figure \@ref(fig:boxplot2).

```{r boxplot2, fig.cap= "A boxplot of Distance from Departing Airports in New York in Different Months"}
ggplot(flights, aes ( origin,distance))+
  geom_boxplot()+
  facet_wrap(~ month)
```

Here boxplot is created for the `distance` variables by `origin` and `month` variables
